Experience Report: Haskell in Computational Biology

### 2012-07-02 mrfy_experience_report.tex
clarify the viterbi-only benchmark
Tried to shorten a few things.

Even with this trick,
MRFy's implementation of Viterbi's algorithm is much
slower than the C++ version implemented in \mrfy\'s predecessor,
SMURF.
For example, on a micro-benchmark that searches for a
structural motif of 343 nodes in a protein of 2000 amino acids,
\mrfy\ takes 2.32~seconds and
SMURF takes 0.29~seconds.

But \mrfy's job is not to run Viterbi's algorithm on large models;
\mrfy's job is to detect homologies to structures on which Viterbi's
algorithm, and SMURF's more complex algorithm, are unsuited.
SMURF may be faster than \mrfy\ in some cases,
but \mrfy\ can solve problems that SMURF cannot.
\mrfy\ can compute an alignment to a complex, 12-stranded ``beta sandwich''
structure within one minute, while on the same structure, SMURF allocates
over 16GB of memory and does not terminate even after eight ours.
The Markov random field model for this structure contains 252 nodes,
97 of which appear in the 12 beta strands.
The models between beta strands typically have at most 12~nodes.
We used a query sequence of 237 amino acids, but each placement
breaks the sequence into 13 pieces, each of which typically has at most
11~amino acids.
On these smaller problems, performance is good enough that \mrfy\ can
do its job.

We benchmarked \mrfy\ using a model of an ``8-bladed beta propeller,''
which has 343 nodes, of which 178 appear in 40~beta strands.
The models between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a very high degree of parallelism.
And, in Haskell, we~found it easy
to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
Using
\texttt{Control.Parallel}, parallelizing this computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:speedup} shows speedup when
using a model of an ``8-bladed beta propeller'' with 1~to~48 cores.
We can conclude that, at least on this particular system,
a single instance of \mrfy\ scales well to about 12~cores.
However, typical usage of \mrfy\ includes running many searches
in parallel, which gives us a way to use all 48 cores on this system.

@@

GHC's profiler assigns costs not to functions but to ``cost
centers'' \cite{sansom-pj}.
In GHC~7.0, which we used 
for most of MRFy's development, 
cost-center annotations can be added automatically only
to top-level 
functions.
But most of our code is in nested functions.
We~annotated such functions manually,
but the annotations
made our code so ugly that we felt compelled to
remove them as soon as possible.
Without the annotations, however, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose profiler gives useful information
without \emph{any} annotations.
GHC~7.4 provides much
better profiling tools, which eliminate most of these obstacles,
rendering this experience no longer as relevant.

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.
Noah Daniels and Andrew Gallant were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).

>>>>

Even with this trick,
MRFy's implementation of Viterbi's algorithm is much
slower than the C++ version implemented in \mrfy\'s predecessor,
SMURF.
For example, on a micro-benchmark that searches for a
structural motif of 343 nodes in a protein of 2000 amino acids
using only the Viterbi algorithm and no beta-strand information,
\mrfy\ takes 2.32~seconds and
SMURF takes 0.29~seconds.

But \mrfy's job is not to run Viterbi's algorithm on large models;
\mrfy's job is to detect homologies to structures on which Viterbi's
algorithm, and SMURF's more complex algorithm, are unsuited.
SMURF may be faster than \mrfy\ in some cases,
but \mrfy\ can solve problems that SMURF cannot.
We have tested \mrfy\ on a complex, 12-stranded ``beta sandwich''
structure, which contains 252 nodes,
97 of which appear in the 12 beta strands.
% The models between beta strands typically have at most 12~nodes.
% We used a query sequence of 237 amino acids, but each placement
% breaks the sequence into 13 pieces, each of which typically has at most
% 11~amino acids.
\mrfy\ computes an alignment in this instance in under one minute,
while on the same structure, SMURF allocates
over 16GB of memory and does not terminate even after eight ours.

We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller,''
which has 343 nodes, of which 178 appear in 40~beta strands.
The models between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a very high degree of parallelism.
And, in Haskell, we~found it easy
to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
Using
\texttt{Control.Parallel}, parallelizing this computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:speedup} shows speedup when
using a model of an ``8-bladed beta propeller'' with 1~to~48 cores.
We can conclude that, at least on this particular system,
a single instance of \mrfy\ scales well to about 12~cores.
However, typical usage of \mrfy\ includes running many searches
in parallel, which gives us a way to use all 48 cores on this system.

@@

GHC's profiler assigns costs not to functions but to ``cost
centers'' \cite{sansom-pj}.
In GHC~7.0, which we used 
for most of MRFy's development, 
cost-center annotations can be added automatically only
to top-level 
functions.
But most of our code is in nested functions.
We~annotated such functions manually,
but the annotations
made our code so ugly that we felt compelled to
remove them as soon as possible.
Without the annotations, however, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose profiler gives useful information
without \emph{any} annotations.
GHC~7.4 provides much
better profiling tools, which eliminate most of these obstacles,
rendering this experience no longer as relevant.
\remark{given that this experience is no longers as relevant,
and we're tight on space, should we cut this entire paragraph
about profiling? --ND}

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.
ND and AG were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).
